Praise the people or praise the place?

### 2025-04-15 paper.qmd
update Kappa'

---
title: "Paper 4: Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-scraper> and <https://github.com/j-jayes/svensk-industrikalender>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |
  Abstract
date: 12 January, 2024
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
execute: 
  echo: false
  warning: false
  message: false
bibliography: bibliography.bib
---

## Introduction

Economic history as a discipline is set to benefit in at least three ways from the nascent AI revolution. Novel sources of data become available as new tools make them readable to machines, analysis in new ways is possible with new kinds algorithms, and the time-cost of asking certain kinds of questions decreases, opening up new avenues for research. In this paper, I make an attempt to leverage these benefits to answer the question, 'who were the high skilled workers in electricity related occupations in Sweden, and where did they come from?'. I detail the process of structuring and analyzing two novel data sources, a set of biographical dictionaries and an industrial catalogue, in order to answer this question.

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayes_power_2024] revealed how the advent of electricity in certain parishes led to positive economic outcomes: a boost in income levels, reduced inequality, and the maintenance of employment levels despite the advent of labor-saving technology. A particularly striking observation was the tendency of workers in these early electrified parishes to remain in their birthplaces, hinting at a newfound economic vibrancy stemming from the income spillovers into sectors not affected by the new technology.

Building on these insights, the present paper delves deeper into the human aspect of this technological revolution. It poses a critical question: Who were the key figures driving this change? Were they local talents nurtured by the opportunities at hand, or did they represent a wave of skilled individuals drawn from afar, lured by the pioneering spirit of these early electrified areas?

To answer this, the investigation leverages two novel and rich data sources. The first, *Vem är Vem*, is a comprehensive set of biographical dictionaries containing the profiles of 75,000 notable Swedes active between 1945 and 1968. The second, the "Svensk Industrikalender" or Swedish Industrial Calendar of 1947, offers an exhaustive catalogue of industrial firms, detailing their activities, workforce, and financial metrics.

I digitize and structure these sources in order to analyze the changing patterns of the Swedish labour market in the middle of the 20th century in light of electrification. The findings challenge my prior expectations. Contrary to the belief that local talent pools predominantly fueled the technological boom, I observe a pattern of geographical mobility among the highly educated and skilled professionals in electricity-related fields. These individuals, pivotal to overseeing and advancing the electricity sector, often sought education and opportunities far from their origins. This suggests a bifurcated labor market: local talent predominantly filled the burgeoning middle-skilled roles within the electricity sector, while the top-tier skilled professionals were more transient, moving towards educational and occupational opportunities. This paper explores the implications of this labor market structure for the economic development patterns witnessed during Sweden's second Industrial Revolution.

These findings, tentative as they are, have real world value. As we seek to understand what drove the dynamism during the age of electrification in Sweden, we are better equipped to shape policy today that seeks to revitalize deindustrializing areas across the developed world and help the developing world harness new technologies for sustainable growth. In addition, the methodologies employed to structure and analyze archival data can provide a template for future research using similar materials.

The paper is laid out as follows: the current research question is placed in context, the sources are explained, followed by their digitization and structuring process. I then discuss the classification task involved in assigning engineers to a sector based on their occupational trajectories. I then lay out some descriptive statistics and tentative findings regarding the patterns of movement for the high skilled electricity related workers, compared with other professionals I observe in the biographical dictionaries.

## Related Literature

@@

Several recent papers have made use of biographic data in innovative ways. @ford2023not use biographies of high school graduates compiled at the time of their school reunions to create a far richer measure of human capital than the conventional measure, number of years of schooling, alone. Titled, "Not the Best Fillers in of Forms? The Danish and Norwegian Graduate Biographies and 'Upper Tail Knowledge'", the authors explain that these biographies are "mini-CVs", containing information about the school leaver's grades, their occupational trajectory, and their family background. These are used to create an innovative approach to measuring `upper tail knowledge`.

@Nekoei2020Herstory titled "Herstory: the rise of self-made women" analyzes the historical prominence of self-made women using a specially created database. This database, formed by applying machine learning to Wikidata and Wikipedia, catalogues notable individuals throughout history, highlighting details like occupation and family background. Their unique approach reveals a significant increase in the number of prominent women, especially those who achieved success independently of their family connections, across various fields, starting with literature, since the 17th century. This research provides a fresh perspective on women's historical achievements and roles.

@@

In this paper, I structure biographic data about elite individuals that is similar in structure to @ford2023not, and use the career trajectories of these individuals to better understand the contribution of educated workers to the adoption of electricity across Sweden in the 20th century. The differentiator in this case is the scale - I capture 75,000 individuals in Sweden in an automated manner, or about one percent of the population of the country at the time.

@@

The historical adaptability of labor markets to technological change is well-documented. In their study of the U.S. labor market's response to the automation of telephone operation, Feigenbaum and Gross @Feigenbaum2020 demonstrate how technological displacement in one sector led to increased demand in others, suggesting an inherent resilience in labor markets. This finding is particularly pertinent to this exploration of Sweden's electrification, as it indicates a potential for both displacement and opportunity in the face of technological change.

Claudia Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to this study [-@Goldin1994; -@Goldin1998]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

The impact of the Digital Revolution on labor markets, as reviewed in the Oxford Review of Economic Policy, is also salient to our study [@Adams2018; @Goos2018]. These articles underscore the emergence of job polarization and the crucial role of policy interventions in ensuring equitable benefits from technological advancements. This perspective is instrumental in understanding the differential impacts of electrification in Sweden, especially in terms of job creation and labor market segmentation.

Moretti’s exploration of the geographical clustering of talent and innovation in "The New Geography of Jobs" provides a crucial perspective on the spatial dynamics of economic development [@Moretti2012]. His findings about the importance of local ecosystems in fostering innovation and economic vitality resonate with our investigation of how early electrification in Swedish parishes influenced the distribution and impact of skilled labor. His concern, that gains to productivity are eaten up by increased cost of living (primarily though housing costs) when constraints prevail, is not evidenced in the first half of the 20th century in Sweden. However, his example of Silicon Valley -- where high productivity and attractive jobs draw in people with high levels of skill, raising property prices - is becoming more concerning in today's relatively housing scarce urban centers.

New technologies require new skills. Mokyr's research provides insights into the importance of both artisans and engineers in the progression of the Industrial Revolution. His studies underscore the synergistic relationship between theoretical knowledge and practical expertise, essential in driving technological innovation and economic progress [@Mokyr2017]. In his examination of the socio-economic elites of early modern Europe, Mokyr explains how their education and exposure to new ideas and sciences were pivotal in fostering various intellectual and technological advancements. This educated elite, through their changing culture and institutions, played a crucial role in creating an environment conducive to innovation [@Mokyr2017Journal].

Not every innovator needs higher education. Mokyr's perspective is crucial in understanding the dynamics of technological development and economic growth, emphasizing the collaborative efforts between well-educated scientists and highly skilled artisans. This interplay highlights the importance of practical skills, theoretical knowledge, and their combined impact on technological progress. For example, figures like metallurgist Henry Cort, who collaborated with scientists despite lacking formal scientific training, exemplify the productive synergy between different forms of expertise in this era [@Mokyr2017Journal].

@@

Relating to the former, @mullainathan2017machine wrote a pathbreaking article that documents the use of machine learning as part of an economists toolkit in the context of prediction problems, differentiating it from traditional parameter estimation. The authors explain the use of new data types like satellite images and text, as well as machine learning's role in policy, estimation, and testing economic theories.

Some interesting papers that incorporate prediction and classification problems include @Bandiera2020ceo, who classify CEO behavior by collecting high-frequency diary information and then use a machine learning algorithm to classify CEOs into 'leaders' and 'managers' by the content of their meetings. @koschnick2023breaking uses a machine learning topic model to classify each paper by the universe of all students at English universities in the seventeenth and early eighteenth century to calculate a measure of how innovative the paper was; how it differed from the papers before it in the field and how similar the papers afterwards were. @DahlVedel2024 in a paper titled "Breaking the HISCO Barrier: AI and Occupational Data Standardization" apply a neural network to the task of classifying an occupational description and benchmark their results against human labelling to show that the neural network achieved comparable accuracy with human labelling and involved an order of magnitude fewer human hours.

'Big data' papers in economic history now abound, as surveyed in @gutmann2018big in a review titled *'Big Data' in Economic History*. Many of these papers construct and use high quality register or census-like data on individual outcomes. Notable examples include the Longitudinal, Intergenerational Family Electronic Micro-Database Project by Martha Bailey which focuses on family histories to understand long-term economic trends. These histories are collected from various census-like sources and innovative ML tools are used to construct the longitudinal links. Similarly, "The Making of Modern America: Migratory Flows in the Age of Mass Migration" by @bandiera2013making involved the digitization of 24 million records of migrant flows through Ellis Island in New York, and found that measured out-migration rates in the US were double the reported figures in the earliest decades of the 20th century. @eichengreen2021gold uses natural language processing (NLP) to understand the content of a parliamentary committee debate on the gold standard in South Africa. Clark and Cummins' families of England database contains 1.7m marriage records in England from the 19th to the 21st centuries and allows to authors to pry out social dynamics in family formation, as well as geographic sorting between the North and South of England [-@Clark2018BigSort].

There is also a growing literature in which authors lay out the step by step processes required for economic historians to make use of these new tools. A great paper in this vein that bridges the gap between cutting-edge computer science literature and the use cases of applied economists is by @correia2023digitizing. The paper "Digitizing Historical Balance Sheet Data: A Practitioner’s Guide" explores the application of machine learning, particularly Optical Character Recognition (OCR), to digitize large-scale historical economic data. The authors highlight the limitations of off-the-shelf OCR software, mainly due to high error rates, and propose a combination of pre- and post-processing methods to enhance accuracy. They apply these methods to two extensive datasets of balance sheets and introduce "quipucamayoc," a Python package that unifies these techniques.

@amujala2023digitization, in "Digitization and data frames for card index records" explain the entire process through which they digitize and structure loan records from bank cards that contain both machine written and hand written text in varying formats over time. Tha authors lower the barrier to entry for other researchers by explaining their use off the shelf technology from Amazon Web Services. Each step is explained along with tips for successful extraction of handwritten information.

Perhaps the most impressive of these kinds of papers is the *Layout Parser* from Melissa Dell and her lab. The team have produced a python library that can be fine tuned to parse document type, extract information, and use machine learning to correct common errors at the point of data extraction. Dell demonstrates the use of this tool by extracting firm performance data from Japanese reports on yearly firm output which are atypical from the kinds of tables or column text that off the shelf optical character recognition tools have been trained on [-@dell2020large].

@@

## Source material

@@

*Vem är Vem?* is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two regional editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house [@harnesk1945vem]. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods.

The primary intention behind the creation of *Vem är Vem?* was to spotlight individuals who were at the peak of their careers, regardless of their age. This focus extends beyond traditional measures of influence, emphasizing the importance of those in influential positions or notable roles across relatively diverse sectors. As such, it serves as a crucial resource for understanding the professional and personal trajectories of around 75,000 individuals who shaped Swedish society in the mid-20th century.

@@

```{r}
#| label: fig-who-is-who-books
#| fig-cap: "Number of biographies in each volume of 'Vem är Vem?'"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/who-is-who-volumes.png")
```

@@

```{r}
#| label: fig-who-is-who-page
#| fig-cap: "A representative page of Vem är Vem?, highlighting the biography of Karl Gustaf Lund"

@@

@@

### Industrial Catalogue

The *Svensk Industrikalender*, published by Sveriges Industriförbund (Sweden's Industrial Association), is a comprehensive directory of Swedish industrial firms. This calendar was issued annually from 1918 to 2000 and contains information related to Swedish industry. The 1947 edition available on the *Project Runeberg* website was digitized in April 2012, sourced from the Centrum för Näringslivshistoria. The calendar is believed to be under catalog protection but not copyright .

It includes detailed information such as company names, locations, nature of businesses, products, contact details, share capital, number of employees, production values, establishment years, and key personnel including managing directors and board members. This source is valuable for studying the economic and industrial environment of post-war Sweden, providing insights into corporate structures, industry distribution, and business trends of that period.

A representative page is shown in @fig-industry-calender.

```{r}
#| label: fig-industry-calender
#| fig-cap: "A page of the industry catalogue from 1947"
#| fig-width: 6
#| out-width: 6in

knitr::include_graphics("assets/svensk-industrikalender_1.png")
```

The common fields listed for each company entry in the catalogue are as follows:

1. **Company Name**: The name of the company is listed at the beginning of each entry, with an asterisk indicating membership of Sveriges Industriförbundet.

2. **Location/Town**: The town or location of the company, which in this case is `Arboga`.

3. **Description of Business**: A brief description of the company's main activities or products is provided.

4. **Products or Services Offered**: Specific items or services the company provides, such as types of machinery, tools, or materials.

5. **Contact Information**: This typically includes:
   - **Telegraph Address**: Listed as "Telegr.-adr." indicating the address to which telegraphs are to be sent.
   - **Telephone Number**: Listed as "Telefon" followed by the number.
   - **Postal Code**: Mentioned as "Postgiro" or "Postiro" with corresponding numbers.
   - **Bank Account**: Sometimes a bank account number or similar financial information is included.

6. **Management and Key Personnel**: Names and titles of important figures in the company, such as the director (`Verkst. dir.`), board members, or founders.

7. **Financial Information**: Information about the financial aspect of the company, such as capital invested (`Akt.-kap.`) or turnover (`Tillv.-värde`).

8. **Establishment Details**: This includes the year of establishment and sometimes the history or lineage of the company's ownership or major changes.

9. **Address**: The full postal address, which may include a street name or a postbox number, indicated as `Postgiro`.

This type of catalogue was commonly used for business-to-business interactions and can be considered an early form of networking resource, allowing companies to find suppliers, customers, and partners.

I use the industry calendar to uncover geographic clusters of firms that produced goods related to electricity for sale to consumers or to other businesses. I use these geographic clusters to understand the migration patterns among engineers in Sweden.

## Data collection strategy

In order to analyze both the biographical dictionaries and industrial catalogue, we need to bend the text into a machine readable structure. This process is not complicated, but somewhat involved. It includes breaking each component of the source up (e.g. each biography or company record), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

```{r}
#| label: fig-data-collection-process
#| fig-cap: "Data collection process steps"
#| fig-width: 5
#| out-width: 5in

@@

@@

### Intuitive explanation of LLMs contextual 'understanding'

>>>>

---
title: "Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-it>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |  
  This paper investigates the origins, characteristics, and career trajectories of high-skilled workers—engineers and business leaders—central to Sweden's electrification during the mid-20th century. Leveraging a novel dataset constructed from approximately 66,000 unique entries in the Vem är Vem? and Vem är Det? biographical dictionaries, I examine the backgrounds of individuals active in the earliest electrifying regions. While previous research found localized benefits and relative immobility for medium-skilled workers in parishes gaining early access to the "Western Line" power grid (Jayes, Enflo and Molinder, 2025), this study reveals a different dynamic for the high-skilled elite. Probit analyses comparing pioneers working in Western Line parishes before 1930 to their peers show they possessed a distinct profile: they were significantly more likely to have technical education, particularly from KTH (Royal Institute of Technology), specific career experience in the USA, and fathers from agricultural or production/transport/laborer backgrounds, suggesting different selection patterns or pathways to mobility. Furthermore, comparing migrants who moved into these parishes versus locals ("stayers") reveals that migrants were more likely to bring KTH degrees and general overseas experience, indicating that these locations with early access to electricity actively recruited specific external human capital. These findings nuance the "People vs. Place" debate, suggesting early electrification hubs ("the Place") attracted and relied upon high-skilled individuals ("the People") with a specific, internationally-informed technical skillset, blending external talent with the local workforce to drive technological advance.

date: today
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
# format:
#   docx:
#     toc: true
#     number-sections: true
#     highlight-style: github
execute: 
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

## I. Introduction

Economic history as a discipline is set to benefit in at least three ways from the nascent AI revolution. Novel sources of data become available as new tools make them readable to machines, analysis in new ways is possible with new kinds of algorithms, and the time-cost of asking certain kinds of questions decreases, opening up new avenues for research. In this paper, I leverage these benefits to explore a fundamental question concerning Sweden's industrial transformation: 'who were the high-skilled workers in electricity-related occupations, and where did they come from?'. I detail the process of structuring and analyzing novel data sources – extensive biographical dictionaries – to answer this question.

Consider the journey of Ivan Beckius, born in 1895, the son of a farmer in rural Västmanland. His path would lead him far from the farm. After technical school in Malmö in 1915, he earns an engineering degree from Stockholm's prestigious Royal Institute of Technology (KTH) in 1919. By 1920, he's a laboratory engineer at ASEA in Västerås – a city situated squarely on the newly established 'Western Line,' the backbone of Sweden's growing national power grid. Later, Beckius ventured across the Atlantic, spending 1923-1925 designing transformers for the industrial giant General Electric in Pittsfield, Massachusetts, and working briefly on power station construction in New York. He returned to ASEA, but his career continued its international trajectory, later managing transformer calculations and technical operations for ASEA in the UK. Beckius's trajectory—rooted in rural Sweden, forged at elite institutions, shaped by international experience, and central to the electrification effort—epitomizes the questions this paper explores: Who were the highly skilled individuals driving this technological revolution? What were their origins, education, and experiences? And in the interplay between emerging industrial centers and mobile talent, should we 'Praise the People' for their skills or 'Praise the Place' for its ability to attract them?

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors examined the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayesPowerProgressImpact2025] shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" revealed how the advent of electricity along the 'Western Line' led to significant positive economic outcomes for the general population: a boost in income levels, reduced inequality, and sustained employment, particularly benefiting lower-income workers with only primary education. That study also noted a tendency for these medium-skilled workers to remain in their birthplaces, suggesting localized benefits and perhaps limited mobility.

Building on these insights, the present paper shifts focus to the upper tail of the human capital distribution – the highly skilled engineers and business leaders like Beckius, whose expertise was crucial for designing, implementing, and managing the new electrical systems. Did the parishes that electrified first attract or cultivate a distinct type of high-skilled talent? Specifically, I ask: What were the characteristics of engineers and business leaders working in the early-electrifying Western Line parishes before 1930? How did their origins, education, and experiences compare to their high-skilled peers active elsewhere or later? Furthermore, within these hubs, did the locals ("stayers") differ significantly from the migrants who moved in ("movers-in")?

To answer these questions, this paper leverages newly combined and structured data from two comprehensive sets of Swedish biographical dictionaries, Vem är Vem? and Vem är Det?, yielding ~66,000 unique individual biographies from the mid-20th century, from which we identify ~48,000 engineers and business leaders. Digitizing and structuring this rich source using modern computational techniques allows us to analyze the backgrounds and career paths of this elite group in unprecedented detail.

My analysis reveals a complex picture that challenges simple narratives. I confirm the high geographic mobility of this elite group, contrasting sharply with the medium-skilled workers studied previously. More importantly, quantitative analysis shows that the high-skilled individuals working in Western Line parishes before 1930 possessed a distinct profile compared to their peers: they were significantly more likely to come from agricultural or working-class backgrounds, possess technical education (specifically from KTH), and have career experience from the United States. Further analysis comparing migrants and locals within these early hubs indicates that migrants were more likely than locals to bring KTH degrees and general overseas experience, suggesting that these places with early access to electricity actively recruited specific external talent to complement the local workforce.

These findings hold value for understanding the drivers of dynamism during Sweden's electrification and offer insights for contemporary policy aimed at fostering growth through technological change. The methodologies employed also provide a template for leveraging rich archival biographical data.   
The paper is laid out as follows: Section 2 places the research question in the context of the relevant literature. Section 3 details the biographical data sources and the methods used for digitization, structuring, and occupation classification. Section 4 presents the empirical strategy. Section 5 presents the results comparing the characteristics of high-skilled pioneers in early electrifying parishes, first to their peers and then comparing migrants versus locals within that group. Section 6 discusses the implications of these findings for the "People vs. Place" debate and our understanding of high-skilled labor's role in technological transitions. Section 7 concludes.

## II. Context

@@

Several recent papers have made use of biographic data in innovative ways. @ford2023NotBestFillersa use biographies of high school graduates compiled at the time of their school reunions to create a far richer measure of human capital than the conventional measure, number of years of schooling, alone. Titled, "Not the Best Fillers in of Forms? The Danish and Norwegian Graduate Biographies and 'Upper Tail Knowledge'", the authors explain that these biographies are "mini-CVs", containing information about the school leaver's grades, their occupational trajectory, and their family background. These are used to create an innovative approach to measuring upper tail knowledge.

@nekoeiHERSTORYRiseSelfmade2021 titled "Herstory: the rise of self-made women" analyzes the historical prominence of self-made women using a specially created database. This database, formed by applying machine learning to Wikidata and Wikipedia, catalogues notable individuals throughout history, highlighting details like occupation and family background. Their unique approach reveals a significant increase in the number of prominent women, especially those who achieved success independently of their family connections, across various fields, starting with literature, since the 17th century. This research provides a fresh perspective on women's historical achievements and roles.

@@

In this paper, I structure biographic data about elite individuals that is similar in structure to @ford2023NotBestFillersa, and use the career trajectories of these individuals to better understand the contribution of educated workers to the adoption of electricity across Sweden in the 20th century. The differentiator in this case is the scale - I capture 66,000 individuals in Sweden in an automated manner, or about one percent of the population of the country at the time. Also, I don't have access to the grade lists that @ford2023NotBestFillersa do, so I cannot use the same kind of measure of human capital.

@@

The historical adaptability of labor markets to technological change is well-documented. In their study of the U.S. labor market's response to the automation of telephone operation, @feigenbaum2020AnsweringCallAutomationa demonstrate how technological displacement in one sector led to increased demand in others, suggesting an inherent resilience in labor markets. This finding is particularly pertinent to this exploration of Sweden's electrification, as it indicates a potential for both displacement and opportunity in the face of technological change.

Claudia Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to this study [-@goldin1994LaborMarketsTwentieth]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

The impact of the Digital Revolution on labor markets, as reviewed in the Oxford Review of Economic Policy, is also salient to this study [@adams2018TechnologyLabourMarket; @goos2018ImpactTechnologicalProgressa]. These articles underscore the emergence of job polarization and the crucial role of policy interventions in ensuring equitable benefits from technological advancements. This perspective is instrumental in understanding the differential impacts of electrification in Sweden, especially in terms of job creation and labor market segmentation.

Moretti’s exploration of the geographical clustering of talent and innovation in "The New Geography of Jobs" provides a crucial perspective on the spatial dynamics of economic development [@moretti2012NewGeographyJobs]. His findings about the importance of local ecosystems in fostering innovation and economic vitality resonate with the investigation of how early electrification in Swedish parishes influenced the distribution and impact of skilled labor. His concern, that gains to productivity are eaten up by increased cost of living (primarily through housing costs) when constraints prevail, is not evidenced in the first half of the 20th century in Sweden. However, his example of Silicon Valley -- where high productivity and attractive jobs draw in people with high levels of skill, raising property prices - is becoming more concerning in today's relatively housing scarce urban centers.

New technologies require new skills. Mokyr's research provides insights into the importance of both artisans and engineers in the progression of the Industrial Revolution. His studies underscore the synergistic relationship between theoretical knowledge and practical expertise, essential in driving technological innovation and economic progress [@mokyr2017CultureGrowthOrigins]. In his examination of the socio-economic elites of early modern Europe, Mokyr explains how their education and exposure to new ideas and sciences were pivotal in fostering various intellectual and technological advancements. This educated elite, through their changing culture and institutions, played a crucial role in creating an environment conducive to innovation [@mokyr2018BottomupTopdownOriginsa].

Not every innovator needs higher education. Mokyr's perspective is crucial in understanding the dynamics of technological development and economic growth, emphasizing the collaborative efforts between well-educated scientists and highly skilled artisans. This interplay highlights the importance of practical skills, theoretical knowledge, and their combined impact on technological progress. For example, figures like metallurgist Henry Cort, who collaborated with scientists despite lacking formal scientific training, exemplify the productive synergy between different forms of expertise in this era [@mokyr2018BottomupTopdownOriginsa].

In Sweden, Andersson and Molinder (@andersson2024DidCitiesIncreasea) investigated how migration influenced skill acquisition during industrialization (1880s–1930s). Using data from the Historical Swedish Population Panel (HISP) and linked censuses, they found that rural-urban migrants significantly increased their occupational income — a proxy for skills — after moving to cities. The impact of cities on skills had both static and dynamic effects. The static effect came from immediate access to higher-skilled jobs, while the dynamic effect reflected skill accumulation over time due to urban learning environments. Migrants with below-median initial incomes benefited the most, highlighting urban areas as drivers of upward mobility. Surprisingly, skill gains were similar across city sizes, including Stockholm and smaller urban centers. This suggests that the benefits of urbanization during Sweden's industrialization stemmed from the general characteristics of urban labor markets rather than city scale.

Heikkuri and Prado (@heikkuri2024TechnologicalChangeSkillsa) examined the impact of electrification on Swedish industry between 1913 and 1926. Using establishment-level data from industrial surveys, they found a positive correlation between electricity adoption, higher employment, and increased wages. Establishments that adopted electricity grew more rapidly in both employment and wages compared to those that did not, highlighting electrification as a key driver of economic advancement.  While direct evidence on skills is lacking, Heikkuri observed occupational shifts from farm laborers to factory workers in electrified parishes. Although both occupations were considered unskilled, factory workers earned significantly higher wages. This aligns with the findings in @jayesPowerProgressImpact2025 and suggests that electrification influenced labor reallocation and improved earnings for unskilled workers. However, Heikkuri and Prado also caution that factors such as collective bargaining agreements may have played a role, underscoring the need for further research to determine electrification's impact on skill demand.

@@

Relating to the former, @mullainathan2017MachineLearningApplied wrote a pathbreaking article that documents the use of machine learning as part of an economists toolkit in the context of prediction problems, differentiating it from traditional parameter estimation. The authors explain the use of new data types like satellite images and text, as well as machine learning's role in policy, estimation, and testing economic theories.

Some interesting papers that incorporate prediction and classification problems include @bandiera2020CEOBehaviorFirm, who classify CEO behavior by collecting high-frequency diary information and then use a machine learning algorithm to classify CEOs into 'leaders' and 'managers' by the content of their meetings. @koschnick2024BreakingTraditionTeacherstudent uses a machine learning topic model to classify each paper by the universe of all students at English universities in the seventeenth and early eighteenth century to calculate a measure of how innovative the paper was; how it differed from the papers before it in the field and how similar the papers afterwards were. @dahl2024BreakingHISCOBarriera in a paper titled "Breaking the HISCO Barrier: AI and Occupational Data Standardization" apply a neural network to the task of classifying an occupational description and benchmark their results against human labelling to show that the neural network achieved comparable accuracy with human labelling and involved an order of magnitude fewer human hours. I make use of this tool in the classification of occupations in this paper.

'Big data' papers in economic history now abound, as surveyed in @gutmann2018BigDataEconomic in a review titled *'Big Data' in Economic History*. Many of these papers construct and use high quality register or census-like data on individual outcomes. Notable examples include the Longitudinal, Intergenerational Family Electronic Micro-Database Project by Martha Bailey which focuses on family histories to understand long-term economic trends. These histories are collected from various census-like sources and innovative ML tools are used to construct the longitudinal links. Similarly, "The Making of Modern America: Migratory Flows in the Age of Mass Migration" by @bandiera2013MakingModernAmerica involved the digitization of 24 million records of migrant flows through Ellis Island in New York, and found that measured out-migration rates in the US were double the reported figures in the earliest decades of the 20th century. @eichengreen2021GoldSouthAfricas uses natural language processing (NLP) to understand the content of a parliamentary committee debate on the gold standard in South Africa. Clark and Cummins' families of England database contains 1.7m marriage records in England from the 19th to the 21st centuries and allows to authors to pry out social dynamics in family formation, as well as geographic sorting between the North and South of England [-@clark2018BigSortDecline].

There is also a growing literature in which authors lay out the step by step processes required for economic historians to make use of these new tools. A great paper in this vein that bridges the gap between cutting-edge computer science literature and the use cases of applied economists is by @correia2023DigitizingHistoricalBalancea. The paper "Digitizing Historical Balance Sheet Data: A Practitioner’s Guide" explores the application of machine learning, particularly Optical Character Recognition (OCR), to digitize large-scale historical economic data. The authors highlight the limitations of off-the-shelf OCR software, mainly due to high error rates, and propose a combination of pre- and post-processing methods to enhance accuracy. They apply these methods to two extensive datasets of balance sheets and introduce "quipucamayoc," a Python package that unifies these techniques.

@amujala2023DigitizationDataFrames, in "Digitization and data frames for card index records" explain the entire process through which they digitize and structure loan records from bank cards that contain both machine written and hand written text in varying formats over time. Tha authors lower the barrier to entry for other researchers by explaining their use off-the-shelf technology from Amazon Web Services. Each step is explained along with tips for successful extraction of handwritten information.

Perhaps the most impressive of these kinds of papers is the *Layout Parser* from Melissa Dell and her lab. The team have produced a python library that can be fine tuned to parse document type, extract information, and use machine learning to correct common errors at the point of data extraction. Dell demonstrates the use of this tool by extracting firm performance data from Japanese reports on yearly firm output which are atypical from the kinds of tables or column text that off-the-shelf optical character recognition tools have been trained on [-@dell2020LargeDatasetHistorical].

@@

## III. Materials and Methods

@@

*Vem är Vem?* is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two regional editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house [@harnesky2013VemArVem]. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods.

The primary intention behind the creation of *Vem är Vem?* was to spotlight individuals who were at the peak of their careers, regardless of their age. This focus extends beyond traditional measures of influence, emphasizing the importance of those in influential positions or notable roles across relatively diverse sectors. As such, it serves as a crucial resource for understanding the professional and personal trajectories of around 66,000 individuals who shaped Swedish society in the mid-20th century.

@@

:::{#fig-who-is-who-books}

```{r}
#| fig-width: 6
#| out-width: 6in
#| column: page
#| fig-align: center

knitr::include_graphics("assets/combined_biographies_plot.png")
```

Source: *Vem är Vem?* and *Vem är Det?* (1945-1968) digitized by @harnesky2013VemArVem and available at [Projekt Runeberg](https://runeberg.org/vemarvem/).

Number of biographies in each volume of 'Vem är Vem?' and 'Vem är Det?'

:::

@@

:::{#fig-who-is-who-page}

```{r}

@@

Source: [Vem är Vem? Götalandsdelen utom Skåne 1948](https://runeberg.org/vemarvem/gota48/0634.html) page 634. Red box added by author.

A representative page of Vem är Vem?, highlighting the biography of Karl Gustaf Lund

:::

@@

### Source criticism

Obviously, the biographical dictionaries are not without their limitations. The criteria for inclusion in the dictionaries was somewhat vague, and individuals could opt in to being included for a nominal fee. It is not a representative sample of the Swedish population at the time. However, there should be some way to assess the representativeness of the biographical dictionaries. One way to do this is to compare the individuals in the biographical dictionaries to a different source of biographical information.

To assess the representativeness of the biographical dictionaries, I sampled a random selection of individuals from these dictionaries and manually checked whether they appear in the Swedish Biographical Lexicon (SBL).

The Swedish Biographical Lexicon (SBL), a personal history lexicon first published in 1917, now (as of spring 2024) includes entries from A to Södersten. It is a general inventory based on primary sources documenting important figures in Swedish society since about the 1500s, comprising approximately 27,200 pages with detailed data on individuals. The SBL aims to include individuals deemed significant by its editors, relying on broad and first-hand sources, without self-selection bias. This approach contrasts with the Vem är Vem? biographical dictionaries, where inclusion was voluntary, with criteria described as "somewhat vague," and individuals could pay to be featured. As a result, Vem är Vem? likely overrepresents those with a certain social standing and willingness to invest in being included, potentially limiting its representativeness of the wider society at the time.

I sampled 200 individuals from Vem är Vem? and found that only 38 (19%) were also present in the SBL, indicating that while Vem är Vem? contains a significant number of entries (about 66,000 individuals, representing roughly 1% of the Swedish population in the mid-20th century), it might overemphasize individuals of higher social status.

The level of detail provided in Vem är Vem? entries is inconsistent. Some individuals have extensive information on their careers, travels, publications, and organizational memberships, whereas others have only basic biographical information. This variability complicates the analysis of certain aspects of individuals' lives, particularly when attempting to classify career trajectories or sectors of work. However, while the amount of information might vary, we can clearly indicate when information is missing for individuals we analyze, and being transparent about these gaps is useful in understanding the limitations of my data.

### Data collection strategy

Analyzing the biographical dictionaries required transforming the unstructured biographical text into a structured, machine-readable format suitable for quantitative analysis. This process is not complicated, but somewhat involved. It includes breaking each component of the source up (e.g. each biography), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

:::{#fig-data-collection-process}

```{r}
#| fig-width: 5
#| out-width: 5in

@@

Source: Author's own.

Illustration of data collection process

:::

@@
